---
title: Week 17. Workbook introduction to mixed-effects models
subtitle: Written by Rob Davies
order: 9
bibliography: references.bib
csl: psychological-bulletin.csl
---

# Week 17 Mixed-effects models workbook overview {#sec-mixed-effects-workbook-overview}

Welcome to your overview of the work we will do together in **Week 17**.

Many Psychologists conduct studies with **repeated-measures designs** where the experimenter presents a sample of multiple stimuli, for response, to each participant in a sample of multiple participants.
Studies with repeated-measures designs will produce data with a structure that requires the use of mixed-effects models.

We are going to learn about this kind of data, and build on the analysis methods we learned about in the [Conceptual introduction to multilevel data](../part2/01-multilevel.qmd#sec-intro-multilevel) and the [workbook introduction to multilevel data](../part2/01-multilevel-workbook.qmd#sec-multilevel-workbook-overview) . 

We use the same procedure we did for multilevel data but with **one significant change** which we shall explore and seek to understand in-depth.

## Targets {#sec-mixed-effects-workbook-targets}

Our learning objectives again include the development of both concepts and skills.

1.  **skills** -- practice how to tidy experimental data for mixed-effects analysis.
2.  **concepts** -- begin to develop an understanding of crossed random effects of participants and stimuli.
3.  **skills and concepts** -- practice fitting linear mixed-effects models incorporating random effects of participants and stimuli.

## Learning resources {#sec-mixed-effects-workbook-resources}

You will see, next, the lectures we share to explain the concepts you will learn about, and the practical data analysis skills you will develop. Then you will see information about the practical materials you can use to build and practise your skills.

Every week, you will learn best if you *first* watch the lectures *then* do the practical exercises.

::: callout-tip
### Linked resources

To help your learning, you can read about the ideas and the practical coding required for analyses in the chapters I wrote for this course.

- The [chapter: Introduction to linear mixed-effects models](../part2/02-mixed.qmd#sec-intro-mixed)
:::

### Lectures {#sec-mixed-effects-workbook-lectures}

The lecture materials for this week are presented in three short parts.

Click on a link and your browser should open a tab showing the *Panopto* video for the lecture part.

1. Part 1 (17 minutes) **Mixed-effects models**: Repeated measures data and the what, when and how of mixed-effects models.

```{=html}
<iframe src="https://lancaster.cloud.panopto.eu/Panopto/Pages/Embed.aspx?id=2fa60b12-2eca-4412-8672-acd60104da1d&autoplay=false&offerviewer=true&showtitle=true&showbrand=true&captions=false&interactivity=all" height="405" width="720" style="border: 1px solid #464646;" allowfullscreen allow="autoplay" aria-label="Panopto Embedded Video Player" aria-description="02-mixed-1-of-3" ></iframe>
```

2. Part 2 (21 minutes): Working with an example from a study with a repeated measures design, recognizing how people or how stimuli vary, and identifying the motivation for using mixed-effects models, compared to alternative methods.

```{=html}
<iframe src="https://lancaster.cloud.panopto.eu/Panopto/Pages/Embed.aspx?id=7f31cef0-1e39-4712-b2a1-acd601195c78&autoplay=false&offerviewer=true&showtitle=true&showbrand=true&captions=false&interactivity=all" height="405" width="720" style="border: 1px solid #464646;" allowfullscreen allow="autoplay" aria-label="Panopto Embedded Video Player" aria-description="02-mixed-2-of-3" ></iframe>
```

3. Part 3 (15 minutes): What mixed-effects models can do for us, how we can describe the logic, how we code mixed-effects models, what the bits of code do, what the results look like, what the results tell us, and how we should report our models and our findings.

```{=html}
<iframe src="https://lancaster.cloud.panopto.eu/Panopto/Pages/Embed.aspx?id=58ab0c0c-37be-491f-a9e5-acd6011c38dd&autoplay=false&offerviewer=true&showtitle=true&showbrand=true&captions=false&interactivity=all" height="405" width="720" style="border: 1px solid #464646;" allowfullscreen allow="autoplay" aria-label="Panopto Embedded Video Player" aria-description="02-mixed-3-of-3" ></iframe>
```

### Lecture slides {#sec-mixed-effects-workbook-slides}

::: callout-tip
## Download the lecture slides
The slides presented in the videos can be downloaded here:

- [402-week-18-LME-2.pdf](files/402-week-18-LME-2.pdf): high resolution .pdf, exactly as delivered [6 MB];
- [402-week-18-LME-2_1pp-lo.pdf](files/402-week-18-LME-2_1pp-lo.pdf): lower resolution .pdf, printable version, one-slide-per-page [about 900 KB];
- [402-week-18-LME-2_6pp-lo.pdf](files/402-week-18-LME-2_6pp-lo.pdf): lower resolution .pdf, printable version, six-slides-per-page [about 900 KB].

The high resolution version is the version delivered for the lecture recordings. Because the images are produced to be high resolution, the file size is quite big (6 MB) so, to make the slides easier to download, I produced low resolution versions: 1pp and 6pp. These should be easier to download and print out if that is what you want to do.
:::

### Practical materials: data and R-Studio {#sec-mixed-effects-workbook-practical}

we will be working with the **CP reading study** dataset. 
CP tested 62 children (aged 116-151 months) on reading aloud in English. 
In the experimental reading task, she presented 160 words as stimuli. 
The same 160 words were presented to all children. 
The words were presented one at a time on a computer screen. 
Each time a word was shown, each child had to read the word out loud and their response was recorded. 
Thus, the CP reading study dataset comprised observations about the responses made by 62 children to 160 words.

You can read more about these data in the [chapter: Introduction to linear mixed-effects models](../part2/02-mixed.qmd#sec-intro-mixed).

The critical features of the study are that:

-   We have an outcome measure -- the reading response -- observed multiple times.

1.  We have *multiple responses recorded for each participant*: they make one response to each stimulus (here, each stimulus word), for the multiple stimuli that they see in the experimental reading task.
2.  And we have *multiple responses recorded for each stimulus*: one response is made to each stimulus by each participant, for all the participants who completed the task, in a sample of multiple participants.

The presence of these features is the reason why we need to use mixed-effects models in our analysis.

::: callout-important
**Get the data**: get the data file and the .R script you can use to do the exercises that will support your learning.

- You can download the files folder for this chapter by clicking on the link [02-mixed.zip](files/02-mixed.zip).
:::

The practical materials folder includes data files and an `.R` script:

-   `CP study word naming rt 180211.dat`
-   `CP study word naming acc 180211.dat`
-   `words.items.5 120714 150916.csv`
-   `all.subjects 110614-050316-290518.csv`
-   `long.all.noNAs.csv`
-   `402-02-mixed-workbook.R` the workbook you will need to do the practical exercises.

The `.dat` files are *tab delimited* files holding behavioural data: the latency or reaction time `rt` (in milliseconds) and the accuracy `acc` of response made by each participant to each stimulus.

The `.csv` files are *comma separated values* files.
The `words.items` file holds information about the 160 stimulus words presented in the experimental reading (word naming) task. 
The `all.subjects` file holds information about the 62 participants who volunteered to take part in the experiment.

In the following, I will describe a series of steps through which we get the data ready for analysis. However, as we shall see, you can avoid these steps by using the pre-tidied dataset:

-   `long.all.noNAs.csv`

The data files are collected together through the steps set out in the .R script:

-   `402-02-mixed-workbook.R`.

::: callout-important
You can access the sign-in page for [R-Studio Server here](https://psy-rstudio.lancaster.ac.uk/auth-sign-in?appUri=%2F)
:::

#### Practical materials guide {#sec-mixed-effects-workbook-practical-guide}

Here, our learning targets are:

- Practice how to tidy experimental data for mixed-effects analysis.
- Begin to develop an understanding of crossed random effects of subjects and stimuli.
- Practice fitting linear mixed-effects models incorporating random effects of subjects and stimuli.

The aims of the lab session work are:

- Get practice running the code required to tidy and prepare experimental data for analysis.
- Exercise skills by varying model code -- changing variables, changing options -- so that you can see how the code works.
- Use the opportunity to reflect on and evaluate results -- so that we can support growth in development of understanding of main ideas.
- Optional -- get practice running the code: so that you can reproduce the figures and results from the lecture and in the book chapter.

#### The practical exercises {#sec-mixed-effects-workbook-practical-workbook}

Now you will progress through a series of tasks, and challenges, to aid your learning.

::: callout-warning
We will work with the data files:

-   `CP study word naming rt 180211.dat`
-   `CP study word naming acc 180211.dat`
-   `words.items.5 120714 150916.csv`
-   `all.subjects 110614-050316-290518.csv`
-   `long.all.noNAs.csv`
:::

We again split the steps into into parts, tasks and questions.

We are going to work through the following workflow steps: **each step is labelled as a practical part**.

1. Set-up
2. Load the data
3. Inspect the data
4. Visualize relationships
5. Analyze relationships using linear models
6. Visualize relationships for each class
7. Mixed-effects analysis
8. Optional exercises: extensions

In the following, we will guide you through the tasks and questions step by step.

::: callout-important
An answers version of the workbook will be provided after the practical class.
:::

#### Practical Part 1: Set-up

To begin, we set up our environment in R.

##### Practical Task 1 -- Run code to load relevant libraries

Use the `library()` function to make the functions we need available to you.

:::{.callout-tip collapse="true"}
## Code
```{r}
library(broom)
library(gridExtra)
library(lme4)
library(tidyverse)
```
:::

#### Practical Part 2: Load the data

##### Practical Task 2 -- Read in the data file we will be using

The data file is called:

- `BAFACALO_DATASET.RData`

Use the `load()` function to read the data files into R:

:::{.callout-tip collapse="true"}
## Code
```{r}
load("BAFACALO_DATASET.RData")
```
:::

Get summary statistics to inspect what you have got.

:::{.callout-tip collapse="true"}
## Code
```{r}
summary(BAFACALO_DATASET)
```
:::

#### Practical Part 3: Tidy the data

##### Practical Task 3 -- Tidy the data

Tidying the data involves a number of tasks, some essential and some things we do for our convenience.

We start by selecting the variables we want: 

- `class_number`, `participant_id`, `portuguese`, `english`, `math`, `physics`

:::{.callout-tip collapse="true"}
## Hint
- We use the `{dplyr}` function `select()` to pick just the variables needed.
- We create a new data-set (call it what you like), from the original, to hold just the selection of variables we want to use in the analysis steps, following.
:::

:::{.callout-tip collapse="true"}
## Code
```{r}
brazil <- BAFACALO_DATASET %>%
                              select(
                              class_number, participant_id,
                              portuguese, english, math, physics
                              )
```
:::

>**Pract.Q.1.** Do you know what you are doing with `select()`? Can you select a different set of variables?

<!-- > Pract.A.1. You can exercise your skill by checking that you can view the `BAFACALO_DATASET` and select any variable you like. Be careful and make sure you go back to select the variables shown in the given code chunk: we will need them later. -->

##### Practical Task 4 -- Take missing values out of the brazil data

There are several rows in the data-set with missing values.
You need to remove these from the data before you get started with any calculations.

:::{.callout-tip collapse="true"}
## Code
```{r}
brazil <- na.omit(brazil)
```
:::

What have you got?

You should always check the results of a tidying operation, to make sure that what you have got matches what you expected to get.

:::{.callout-tip collapse="true"}
## Code
```{r}
summary(brazil)
```
:::

>**Pract.Q.2.** Do you see the difference between the summary of the brazil data shown before and after you run `na.omit()`? What is it?

<!-- > Pract.A.2. You should see that `NA`s are listed in some columns in the data before but not after you run `na.omit()`. -->

##### Practical Task 5 -- Get R to treat a variable as a type object of the kind required -- using the `as.numeric()` or `as.factor()` functions

:::{.callout-tip collapse="true"}
## Code
```{r}
brazil$portuguese <- as.numeric(brazil$portuguese)
brazil$english <- as.numeric(brazil$english)
brazil$math <- as.numeric(brazil$math)
brazil$physics <- as.numeric(brazil$physics)

brazil$class_number <- as.factor(brazil$class_number)
brazil$participant_id <- as.factor(brazil$participant_id)
```
:::

What have you got?

:::{.callout-tip collapse="true"}
## Code
```{r}
summary(brazil)
```
:::

>**Pract.Q.3.** Do you see the difference between the summary of the brazil data shown before and after you run `as.numeric()` or `as.factor()` function lines? What is it?

<!-- > Pract.A.3. You should see that numeric variables like portuguese are listed with summary statistics like the mean after but not before you coerce the variables. -->

##### Practical Task 6 -- Exercise -- experiment with coercion

1. Test out variable type using the `is...()` function for some of the variables
2. Test out coercion -- and its results -- using the `as...()` function for some of the variables
3. Look at the results using `summary()`

#### Practical Part 4: Visualize the relationship

Here, we are interested in the relationship between school grades, for the sample of Brazilian children for whom we have data.

- What is the relationship between portuguese and english grades?

##### Practical Task 7 -- Visualize the relationship between portuguese and english grades using a scatterplot

This should be a revision exercise for you.

- Do you notice any changes in the way in which the plotting code is laid out?

:::{.callout-tip collapse="true"}
## Code
```{r}
brazil %>%
  ggplot(aes(x = portuguese, y = english)) +
  geom_point(colour = "black", size = 2.5, alpha = .5) +
  geom_smooth(method = "lm", size = 1.5, se = FALSE, colour = "red") +
  xlab("Portuguese") + ylab("English") + theme_bw() 
```
:::

##### Practical Task 8 -- Exercise -- edit plots

1. Change the x and y variables to math and physics
2. Change the theme from `theme_bw()` to something different
3. Change the appearance of the points, try different colour, shape, size

:::{.callout-tip collapse="true"}
## Hint
Use the ggplot reference documentation to help you make choices:

- <https://ggplot2.tidyverse.org/reference/>

You should be using webpages like the reference page often in your work.
:::

#### Practical Part 5: Visualize the relationship

We move next to analyzing the relationship between variation in grades in different subjects, across the children in our sample.

##### Practical Task 9 -- Analyze the relationship between english and portuguese grades in the brazil data

You should be able to reproduce the results shown in the slides and the book chapter.

:::{.callout-tip collapse="true"}
## Code
```{r}
summary(lm(english ~ portuguese, data = brazil))
```
:::

##### Practical Task 10 -- Exercise -- adapt the `lm()` code to do a different analysis

Change the outcome and predictor variables to math and physics.

:::{.callout-tip collapse="true"}
## Code
```{r}
summary(lm(physics ~ math, data = brazil))
```
:::

>**Pract.Q.4.** What is the estimated coefficient of the "effect" of math ability (grade) on physics grade?

<!-- > Pract.A.4. The summary shows that physics grades are on average 1.3 higher for unit increase in math grade. -->

>**Pract.Q.5.** Draw a scatterplot showing the relationship between math and physics grades. Does the trend you see in the plot reflect the coefficient you see in the linear model summary?

<!-- > Pract.A.5. The plot shows the positive association between math and physics grade also indicated by the estimated coefficient of the math effect. -->

:::{.callout-tip collapse="true"}
## Code
```{r}
brazil %>%
  ggplot(aes(x = math, y = physics)) +
  geom_point(colour = "black", size = 2.5, alpha = .5) +
  geom_smooth(method = "lm", size = 1.5, se = FALSE, colour = "red") +
  theme_bw() 
```
:::

>**Pract.Q.6.** How does the strength of the math-physics relationship compare with the english-portuguese relationship?

<!-- > Pract.A.6. Both the linear model and the plots indicate that the math-physics relationship is much stronger. -->

#### Practical Part 6: Visualize the relationship for each class

##### Practical Task 11 -- Plot the relationship between english and portuguese grades separately for each class using `facet_wrap()`

:::{.callout-tip collapse="true"}
## Code
```{r}
ggplot(data = brazil, aes(x = portuguese, y = english)) +
  geom_point(colour = "darkgrey") + 
  geom_smooth(method = "lm", se = FALSE, colour = "black") +
  xlab("Portuguese") + ylab("English") + theme_bw() +
  scale_x_continuous(breaks=c(25,50,75)) + scale_y_continuous(breaks=c(0,50,100)) +
  facet_wrap(~ class_number) 
```
:::

##### Practical Task 12 -- Exercises to practice your `facet_wrap()` skills

1. Change the x and y variables to math and physics and draw a facetted scatterplot again
2. Experiment with showing the differences between classes in a different way: instead of using `facet_wrap()`, in `aes()` add `colour = class_number`, and remove colour from `geom_point` and `geom_smooth`

:::{.callout-tip collapse="true"}
## Code
```{r}
ggplot(data = brazil, aes(x = math, y = physics)) +
  geom_point(colour = "darkgrey") + 
  geom_smooth(method = "lm", se = FALSE, colour = "black") +
  facet_wrap(~ class_number) 

ggplot(data = brazil, aes(x = math, y = physics, colour = class_number)) +
  geom_point() + geom_smooth(method = "lm", se = FALSE)
```
:::

>**Pract.Q.8.** Evaluate the consistency between classes of the relationship between math and physics grades: what do the plots show? how does this compare with what you see of the relationship between english and portuguese grades?

<!-- > Pract.A.8. The plots show that the relationship between math and physics is very consistent between classes, and more consistent than the relationship between english and portuguese grades appears to be. -->

#### Practical Part 7: Mixed-effects analysis

##### Practical Task 13 -- Run a linear mixed-effects analysis of the relationship between english and portuguese grades using `lmer()`

You should be able to replicate the results shown in the slides and the book chapter.

:::{.callout-tip collapse="true"}
## Code
```{r}
porto.lmer1 <- lmer(english ~ portuguese +
                      (portuguese + 1|class_number),
                      data = brazil)
summary(porto.lmer1)
```
:::

##### Practical Task 14 - Exercise mixed-effects model coding

1. Vary the random effects part of the model, while keeping this bit the same: `lmer(english ~ portuguese + ...)`
2. Change the random effect from `(portuguese + 1 | class_number)` to `(1 | class_number)`: what you are doing is asking R to ignore the differences in the slope of the effect of Portuguese grades.
3. Change the random effect from `(portuguese + 1 | class_number)` to `(portuguese + 0 | class_number)`: what you are doing is asking R to ignore the differences  in the intercept

:::{.callout-tip collapse="true"}
## Code
```{r}
porto.lmer1 <- lmer(english ~ portuguese +
                      (portuguese + 1|class_number),
                    data = brazil)
summary(porto.lmer1)

#

porto.lmer2 <- lmer(english ~ portuguese +
                      (1|class_number),
                    data = brazil)
summary(porto.lmer2)

#

porto.lmer3 <- lmer(english ~ portuguese +
                      (portuguese + 0|class_number),
                    data = brazil)
summary(porto.lmer3)
```
:::

>**Pract.Q.9.** Compare the results of the different versions of the model. Can you identify where the results are different?

<!-- > Pract.A.9. It can be seen that Rhe estimated effect of portuguese varies between the models but the estimate is more similar, around .65, where the random effect is specified as  -->
<!--  `(portuguese + 1|class_number)` or `(portuguese + 0|class_number)`. -->
<!-- > The residual variance term is different between the models. -->
<!-- > Which random effects variances are shown is also different. -->
<!-- > There is a convergence warning for: -->

<!-- `english ~ portuguese + (portuguese + 0 | class_number)` -->

##### Practical Task 15 - Exercise mixed-effects model coding

Change the outcome (from english) and the predictor (from portuguese) -- this is about changing the fixed effect part of the model.

- Note that you will need to change the random effect part as well.

:::{.callout-tip collapse="true"}
## Code
```{r}
porto.lmer1 <- lmer(physics ~ math +
                      (math + 1|class_number),
                    data = brazil)
summary(porto.lmer1)
```
:::

>**Pract.Q.10.** What elements of the model summary stand out for you? It will help to see what you should notice if you compare the math-physics model with the first english-portuguese model.

<!-- > Pract.A.10. You may notice that: -->

<!-- - The summary comes with a fit is singular? warning. -->
<!-- - The variance terms for intercept or the math effect by class number and the residual are very very small: much smaller than for the english-portuguese model. -->

#### Practical Part 8: Extension

##### Practical Task Optional

In the lecture materials, I display a plot showing the estimated intercept and coefficient for  each class, estimated using separate models for different classes.

- Some of you may be interested in how I did that, you can run the following code to see.

1. Use the `{dplyr}` `%>%` syntax to run a model for each class separately, collect together the results into a dataframe.

:::{.callout-tip collapse="true"}
## Code
```{r}
brazlm <- brazil %>% group_by(class_number) %>% do(tidy(lm(english ~ portuguese, data=.)))
brazlm$term <- as.factor(brazlm$term)
```
:::

2. Extract the per-class estimates of the intercepts and the 'portuguese' effect coefficient estimates.

:::{.callout-tip collapse="true"}
## Code
```{r}
brazlmint <- filter(brazlm, term == '(Intercept)')
brazlmport <- filter(brazlm, term == 'portuguese')
```
:::

3. Plot the estimates.

:::{.callout-tip collapse="true"}
## Code
```{r}
pbrazlmint <- ggplot(brazlmint, aes(x = class_number, y = estimate, ymin = estimate - std.error, ymax = estimate + std.error))
pbrazlmint <- pbrazlmint + geom_point(size = 2) + geom_linerange() + theme_bw() 
pbrazlmint <- pbrazlmint + ggtitle("Intercept") + ylab("Estimated coefficient +/- SE") + xlab("Class")
pbrazlmint <- pbrazlmint + theme(axis.title.y = element_text(size = 10), axis.text.x = element_blank(), panel.grid = element_blank())
# pbrazlmint

pbrazlmport <- ggplot(brazlmport, aes(x = class_number, y = estimate, ymin = estimate - std.error, ymax = estimate + std.error))
pbrazlmport <- pbrazlmport + geom_point(size = 2) + geom_linerange() + theme_bw() 
pbrazlmport <- pbrazlmport + ggtitle("Portuguese effect") + ylab("Estimated coefficient +/- SE") + xlab("Class")
pbrazlmport <- pbrazlmport + theme(axis.title.y = element_text(size = 10), axis.text.x = element_blank(), panel.grid = element_blank())
# pbrazlmport

# -- ask R to make a grid:

grid.arrange(pbrazlmint, pbrazlmport,
             ncol = 2)
```
:::

Challenge: Can you change the code to show the estimates for the relationship between physics and math grades?

### The answers

After the practical class, we will reveal the answers that are currently hidden.

The answers version of the webpage will present my answers for questions, and some extra information where that is helpful.

